What We Eval in the Shadows

### 2021-06-27 paper/main.tex
Low-level writing issues

Most dynamic languages provide their users with a facility to
transform unstructured text into executable code and evaluate that
code. We refer to this reflective facility as \eval bowing to its
origins in LISP, all the way back in 1956~\cite{lisp}. \Eval has been
much maligned over the years. In computing lore, it is as close to a
boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
write down the definition of LISP; he was surprised that someone coded
it up and offered it to end-users. Since then, reflective facilities
have been used to parameterize programs over code patterns that can be
provided after the program is written. The presence of such a feature
in a language is a hallmark of dynamism; it is a form of delayed
binding as the behavior of any particular call to \eval will only be
known when the program is run, and that particular call site is
evaluated.

@@

\vspace{2mm}\noindent\emph{The Here and Now.} In this study, we investigate the
usage of \eval in programs written in the R programming language. R is a language
designed by statisticians for applications in data science~\cite{r,R96}. What
makes looking at R after JavaScript interesting is that, while both languages
are dynamic, they are quite different. While one can program in an
object-oriented style in R like in JavaScript, R is primarily a lazy, untyped,
functional language. JavaScript was designed to run untrusted code in a browser,
while R is used for statistical computing on desktops. JavaScript is a general-purpose language used by a vast community of programmers, while R is used for
scientific computing by data scientists and domain experts with, often, limited
programming experience. One can distinguish between library implementers,
developers with some programming experience and a working knowledge of R, and
end-users, who are typically not expert programmers and often have only a
cursory knowledge of the language.\footnote{Consider that R is lazy like
Haskell. We informally surveyed end-users and did not find a single user aware
of this fact. Library developers, on the other hand, know and program
defensively around laziness.} Thus, our goal is to highlight the differences in
usage between JavaScript and R and try to explain those differences in terms of
language features, application domain and programmer experience. Hopefully, some
of our observations will generalize to other languages.

@@

The analysis presented in this has been automated by a pipeline that acquires
packages, extracts scripts, executes them, traces their behavior, and summarizes observations. Figure~\ref{fig:pipeline} shows the main steps along
 with their running time, data size, and  number of elements manipulated.
  Timings are from a cluster of three 2.3GHz Intel Xeon 6140 servers, with 72 cores and 256GB of RAM, and shared OCFS network storage. The pipeline steps are:

@@

\begin{wrapfigure}{r}{5cm} \vspace{-1cm} \hspace*{-12mm}
  \centering
  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}\vspace{-3mm}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. We use the word \emph{site} to an occurrence of a call site to the \eval
function in the source code and \emph{call} to denote an observed invocation of
the \eval function.

@@

\paragraph{Discussion}
\Eval is central to R's implementation as it is omnipresent in the Base library.
That library is, in turn, used by every single package. To avoid Base dominating
the result, we excluded calls to it from the results. CRAN Packages, which are
typically developed by experienced programmers, make regular and varied use of
\eval. They represent our most interesting data set as these packages result
from over 20 years of contributions by thousands of authors. The code in
packages is well maintained and relatively well tested. Finally, Kaggle scripts
are often written by less sophisticated users and likely perform more
straightforward tasks, thus having a lesser need for \eval, and its uses
originate from strings as these are easier to manipulate for end-users.

>>>>

Most dynamic languages provide their users with a facility to
transform unstructured text into executable code and evaluate that
code. We refer to this reflective facility as \eval bowing to its
origins in LISP, all the way back in 1956~\cite{lisp}. \Eval has been
much maligned over the years. In computing lore, it is as close to a
boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
write down the definition of LISP; he was surprised that someone coded
it up and offered it to end users. Since then, reflective facilities
have been used to parameterize programs over code patterns that can be
provided after the program is written. The presence of such a feature
in a language is a hallmark of dynamism; it is a form of delayed
binding as the behavior of any particular call to \eval will only be
known when the program is run, and that particular call site is
evaluated.

@@

\vspace{2mm}\noindent\emph{The Here and Now.} In this study, we investigate the
usage of \eval in programs written in the R programming language. R is a language
designed by statisticians for applications in data science~\cite{r,R96}. What
makes looking at R after JavaScript interesting is that, while both languages
are dynamic, they are quite different. While one can program in an
object-oriented style in R like in JavaScript, R is primarily a lazy, untyped,
functional language. JavaScript was designed to run untrusted code in a browser,
while R is used for statistical computing on desktops. JavaScript is a general-purpose language used by a vast community of programmers, while R is used for
scientific computing by data scientists and domain experts with, often, limited
programming experience. One can distinguish between library implementers,
developers with some programming experience and a working knowledge of R, and
end users, who are typically not expert programmers and often have only a
cursory knowledge of the language.\footnote{Consider that R is lazy like
Haskell. We informally surveyed end users and did not find a single user aware
of this fact. Library developers, on the other hand, know and program
defensively around laziness.} Thus, our goal is to highlight the differences in
usage between JavaScript and R and try to explain those differences in terms of
language features, application domain and programmer experience. Hopefully, some
of our observations will generalize to other languages.

@@

The analysis presented in this has been automated by a pipeline that acquires
packages, extracts scripts, executes them, traces their behavior, and summarizes observations. Figure~\ref{fig:pipeline} shows the main steps along
 with their running time, data size, and  number of elements manipulated.
  Timings are from a cluster of three servers, each with 2.3GHz Intel Xeon 6140 processor with 72 cores and 256GB of RAM, and shared OCFS network storage. The pipeline steps are:

@@

\begin{wrapfigure}{r}{5cm} \vspace{-1cm} \hspace*{-12mm}
  \centering
  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}\vspace{-3mm}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. We use the word \emph{site} to refer to an occurrence of a call site to the \eval
function in the source code and \emph{call} to denote an observed invocation of
the \eval function.

@@

\paragraph{Discussion}
\Eval is central to R's implementation as it is omnipresent in the Base library.
That library is, in turn, used by every single package. To avoid Base dominating
the result, we excluded calls to it from the results. CRAN Packages, which are
typically developed by experienced programmers, make regular and varied use of
\eval. They represent our most interesting data set as these packages result
from over 20 years of contributions by thousands of authors. The code in
packages is well maintained and relatively well tested. Finally, Kaggle scripts
are often written by less sophisticated users and likely perform more
straightforward tasks, thus having a lesser need for \eval, and its uses
originate from strings as these are easier to manipulate for end users.
